---
title: "Enrichment analysis for astrocytes"
output: html_notebook
author: Alexis Felipe Rojas Cruz
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.keep = 'all',
	dpi = 600,
	fig.width = 11,
	fig.height = 8.5
)
```

The purpose of Venn diagrams is to identify shared DEGs with a variety of contrasts and related treatments. For this study, we found that Tibolone has no effect on astrocytes treated with Palmitic Acid. Based on the Venn diagram analyses, it was found the Palmitic Acid induced DEGs listed below to perform the enrichment analysis 



|No.| Treatment	               |Contrast	                                                        |Up DEGs     |Down DEGs    |
|:--|:-------------------------|:------------------------------------------------------------|:----------:|:-----------:|
|1	|Palmitic Acid	           |AstroTIBPA vs AstroVH, AstroPA vs AstroVH                    |366         |373          |




```{r include=FALSE, comment = ""}
if (!("clusterProfiler" %in% rownames(installed.packages()))){
    install.packages("clusterProfiler")
}
if (!("org.Hs.eg.db" %in% rownames(installed.packages()))){
    install.packages("org.Hs.eg.db")
}
if (!("biomaRt" %in% rownames(installed.packages()))){
    install.packages("biomaRt")
}
if (!("GOplot" %in% rownames(installed.packages()))){
    install.packages("GOplot")
}
if (!("RColorBrewer" %in% rownames(installed.packages()))){
    install.packages("RColorBrewer")
}
if (!("tidyverse" %in% rownames(installed.packages()))){
    install.packages("tidyverse")
}
if (!("ggplot2" %in% rownames(installed.packages()))){
    install.packages("ggplot2")
}
if (!("ggrepel" %in% rownames(installed.packages()))){
    install.packages("ggrepel")
}
if (!("ggsci" %in% rownames(installed.packages()))){
    install.packages("ggsci")
}
if (!("cowplot" %in% rownames(installed.packages()))){
    install.packages("cowplot")
}
if (!("DOSE" %in% rownames(installed.packages()))){
    install.packages("DOSE")
}
if (!("ggh4x" %in% rownames(installed.packages()))){
    install.packages("ggh4x")
}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ReactomePA")


library(clusterProfiler)
library(org.Hs.eg.db)
library(biomaRt)
library(GOplot)
library(RColorBrewer)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(ggsci)
library(cowplot)
library(ReactomePA)
library(DOSE)
library(ggh4x)
```

## Loading data

We will employ the intercepts of the shared DEGs across contrasts for the treatment of PA involving up- and down-regulated genes. In addition, filtered genes obtained by differential expression analysis will be included to perform overrepresentation

```{r data}
PA_up <- read.csv("../Overlapping/overlap_DEGs/PA-DEG_up.csv", sep = ",", header = T)
PA_down <- read.csv("../Overlapping/overlap_DEGs/PA-DEG_down.csv", sep = ",", header = T)

univ <- read.csv("../Expression/Annotation/allgenesfilt.csv", sep = ",", header = T)
```

# Over-representation Analyses (ORA)

Over-representation (or enrichment) analysis is a statistical method that determines whether genes from pre-defined sets (ex: those belonging to a specific GO term or pathway) are present more than would be expected (over-represented) in a subset of your data

## Preparing data

```{r preparing, message=FALSE, warning=FALSE}

# Palmitic Acid
PAup_org <- clusterProfiler::bitr(PA_up$PAvsVH_symbol, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db" ) %>%
  dplyr::select( ENTREZID ) %>%
  dplyr::rename( Entrez = ENTREZID ) %>% 
  add_column( Treatment = "PA", Regulation = "Up" )

PAdown_org <- clusterProfiler::bitr(PA_down$PAvsVH_symbol, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db" ) %>%
  dplyr::select( ENTREZID ) %>%
  dplyr::rename( Entrez = ENTREZID ) %>%
  add_column( Treatment = "PA", Regulation = "Down" )

mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", path="/biomart/martservice", dataset="hsapiens_gene_ensembl")
symbols <- getBM(values=univ$gene_id, filters= "ensembl_gene_id", attributes= c("ensembl_gene_id", "hgnc_symbol","entrezgene_id"), mart= mart)

# Universe
univ_enrich <- symbols %>%
  dplyr::rename( gene_id = ensembl_gene_id ) %>%
  dplyr::rename( Entrez = entrezgene_id ) %>%
  merge( univ, ., by="gene_id" ) %>%
  dplyr::select( Entrez ) %>% 
  as.data.frame( . ) %>%
  unique( . )

```

To create a single dataset for our DEGs, we will merge the different data frames


```{r dataset}

astro_enrich <- dplyr::bind_rows( PAup_org, PAdown_org )
head(astro_enrich)
  
```


## GO Terms

`clusterProfiler` was developed for biological theme comparison, and it provides a function, `compareCluster`, to automatically calculate enriched functional categories of each gene clusters. This is an innovative analysis which enables us to compare DEGs by rormula interface _Entrez∼group  or Entrez∼group+othergroup_ and provides differences and similarities among the groups we are studying.

```{r GO_terms}

compareGO = compareCluster(Entrez ~ Treatment+Regulation, data = astro_enrich,  fun ="enrichGO",
                           ont = "ALL", OrgDb = org.Hs.eg.db, pAdjustMethod = "BH", pvalueCutoff  = 0.01,
                           qvalueCutoff  = 0.05, readable = TRUE)


p1 <- dotplot(compareGO, x="Treatment",split = "ONTOLOGY", showCategory = 5) +
  aes(x=fct_relevel(Treatment, c('PA'))) + xlab(NULL) +
  facet_grid(vars(ONTOLOGY), vars(factor(Regulation, levels = c("Up", "Down"))),
  scale = "free", space = "free") +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  force_panelsizes(cols = c(0.01, 0.01)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 28),
  panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22))
p1

ggsave("../Enrichment/enrich_plots/enrichGO.pdf", plot = p1, dpi = 600, width = 9, height = 19)
```


```{r save enrich GO}

GO_PA <- compareGO@compareClusterResult %>% 
  filter( compareGO@compareClusterResult$Treatment == "PA" ) %>%
  arrange( desc(Regulation) )

write.csv(GO_PA, file="../Enrichment/enrich_GO/GO_PA.csv", quote = F, row.names = F)
```

## KEGG Pathway

```{r KEGG}

compareKEGG = compareCluster(Entrez ~ Treatment+Regulation, data = astro_enrich, fun ="enrichKEGG",
                           organism = "hsa", pAdjustMethod = 'BH', pvalueCutoff = 0.01, qvalueCutoff  = 0.05)

p2 <- dotplot(compareKEGG, x="Treatment", showCategory = 7) +
  aes(x=fct_relevel(Treatment, c('PA'))) + xlab(NULL) +
  facet_grid(~factor(Regulation, levels = c("Up", "Down")), scale = "free", space = "free") +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  force_panelsizes(cols = c(0.01, 0.01)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 28),
  panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22))
p2

ggsave("../Enrichment/enrich_plots/enrichKEGG.pdf", plot = p2, dpi = 600, width = 9, height = 6)
```

```{r save enrich KEGG}

KEGG_PA <- compareKEGG@compareClusterResult %>% 
  filter( compareKEGG@compareClusterResult$Treatment == "PA" ) %>%
  arrange( desc(Regulation) )

write.csv(KEGG_PA, file="../Enrichment/enrich_KEGG/KEGG_PA.csv", quote = F, row.names = F)
```

## WikiPathways

```{r Wiki}

compareWP <- compareCluster(Entrez ~ Treatment+Regulation, data = astro_enrich, fun ="enrichWP",
                           organism = "Homo sapiens", pAdjustMethod = 'BH', pvalueCutoff = 0.01, qvalueCutoff  = 0.05)

p3 <- dotplot(compareWP, x="Treatment", showCategory = 7) +
  aes(x=fct_relevel(Treatment, c('PA'))) + xlab(NULL) +
  facet_grid(~factor(Regulation, levels = c("Up", "Down")), scale = "free", space = "free") +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  force_panelsizes(cols = c(0.01, 0.01)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 28),
  panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22))
p3

ggsave("../Enrichment/enrich_plots/enrichWP.pdf", plot = p3, dpi = 600, width = 9, height = 7.5)
```

```{r save enrich Wiki}

Wiki_PA <- compareWP@compareClusterResult %>% 
  filter( compareWP@compareClusterResult$Treatment == "PA" ) %>%
  arrange( desc(Regulation) )

write.csv(Wiki_PA, file="../Enrichment/enrich_Wiki/Wiki_PA.csv", quote = F, row.names = F)
```


## Reactome

```{r Reactome}

compareRE <- compareCluster(Entrez ~ Treatment+Regulation, data = astro_enrich, fun ="enrichPathway",
                           organism = "human", pAdjustMethod = 'BH', pvalueCutoff = 0.01, qvalueCutoff  = 0.05)

p4 <- dotplot(compareRE, x="Treatment", showCategory = 7) +
  aes(x=fct_relevel(Treatment, c('PA'))) + xlab(NULL) +
  facet_grid(~factor(Regulation, levels = c("Up", "Down")), scale = "free", space = "free") +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  force_panelsizes(cols = c(0.01, 0.01)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 28),
  panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22))
p4

ggsave("../Enrichment/enrich_plots/enrichRE.pdf", plot = p4, dpi = 600, width = 9, height = 6)
```

```{r save enrich RE}
RE_PA <- compareRE@compareClusterResult %>% 
  filter( compareRE@compareClusterResult$Treatment == "PA" ) %>%
  arrange( desc(Regulation) )

write.csv(RE_PA, file="../Enrichment/enrich_RE/RE_PA.csv", quote = F, row.names = F)
```


# GoChord


## Palmitic Acid


```{r data preparation PA}

#Upregulated
PAup_process <- GO_PA %>% 
  dplyr::filter( Regulation == "Up") %>%
  dplyr::select( ONTOLOGY, ID, Description, geneID, p.adjust ) %>% 
  dplyr::rename( category = ONTOLOGY ) %>%
  dplyr::rename( term = Description ) %>%
  dplyr::rename( genes = geneID ) %>%
  dplyr::rename( adj_pval = p.adjust ) %>%
  dplyr::mutate( genes = str_replace_all(genes, '/', ',') ) %>% 
  dplyr::slice(1:8) %>% 
  data.frame( . )

PAup_genes <- data.frame(ID=PA_up$PAvsVH_symbol, logFC=PA_up$PAvsVH_logFC, logFC2=PA_up$TiPAvsVH_logFC)

PAup_circ <- circle_dat(PAup_process, PAup_genes)

#Downregulated  
PAdown_process <- GO_PA %>% 
  dplyr::filter( Regulation == "Down") %>%
  dplyr::select( ONTOLOGY, ID, Description, geneID, p.adjust ) %>% 
  dplyr::rename( category = ONTOLOGY ) %>%
  dplyr::rename( term = Description ) %>%
  dplyr::rename( genes = geneID ) %>%
  dplyr::rename( adj_pval = p.adjust ) %>%
  dplyr::mutate( genes = str_replace_all(genes, '/', ',') ) %>% 
  dplyr::slice(1:8) %>% 
  data.frame( . )

PAdown_genes <- data.frame(ID=PA_down$PAvsVH_symbol, logFC=PA_down$PAvsVH_logFC, logFC2=PA_down$TiPAvsVH_logFC)

PAdown_circ <- circle_dat(PAdown_process, PAdown_genes)
```

```{r format PA}
PAup_data <- chord_dat(data = PAup_circ, genes = PAup_genes)
PAdown_data <- chord_dat(data = PAdown_circ, genes = PAdown_genes)
head(PAdown_data)

```

### Upregulated

```{r GOChord PAup}

PAup_chord <- GOChord(data=PAup_data, space=0.1, limit = c(7, 1), gene.order="logFC",
          gene.size=6, gene.space=0.2,border.size=0.15,
          lfc.col=c('red','white'), process.label=8, 
          ribbon.col = brewer.pal(length(PAup_process$term),"Set2"),
          lfc.min=min(PAup_genes$logFC),lfc.max=max(PAup_genes$logFC))

# extracting fold change legend
FCLeg <- get_legend(PAup_chord+guides(size="none")+
    theme(legend.position = c(0.13,-0.25), 
    legend.direction = "horizontal"))

#Format functional terms legend
PAup_chord <-PAup_chord+guides(size = guide_legend("GO Terms", ncol = 1, byrow = T, 
                               override.aes = list(shape = 22, fill = unique(brewer.pal(length(PAup_process$term),"Set2")),
                               size = 8)), fill="none")+
  theme(legend.key.size = unit(0.01, 'cm'), legend.text = element_text(size=14),
        legend.title = element_text(size=18, face="bold"))

#extracting functional terms legend
FuncLeg <- cowplot::get_legend(PAup_chord+guides(color="none")+
     theme(legend.box.margin =unit(c(0,0,0,0), "cm"),
     legend.spacing = unit(c(-0.1,-0.1,-0.1,-0.1), "cm"),
     legend.position="right", legend.text = element_text(size = 22),
     legend.direction = "vertical"))

# final plot
fPA_up <- plot_grid(FCLeg,
    plot_grid(PAup_chord+theme(legend.position = "none", 
    plot.margin = unit(c(0.1,0.1,0,0), "cm")),FuncLeg, 
    ncol=2, rel_widths = c(0.64,0.36)),
    nrow=2, rel_heights = c(0.07,0.93))


ggsave("../Enrichment/GOChord/PA_up.pdf", plot = fPA_up, dpi = 600, width = 16, height = 11)
```

### Downregulated

```{r GOChord PAdown}

PAdown_chord <- GOChord(data=PAdown_data, space=0.1, limit = c(4.5, 1), gene.order="logFC",
          gene.size=6, gene.space=0.2,border.size=0.15,
          lfc.col=c('white','blue'), process.label=8, 
          ribbon.col = brewer.pal(length(PAdown_process$term),"Set2"),
          lfc.min=min(PAdown_genes$logFC),lfc.max=max(PAdown_genes$logFC))

# extracting fold change legend
FCLeg <- get_legend(PAdown_chord+guides(size="none")+
    theme(legend.position = c(0.13,-0.25), 
    legend.direction = "horizontal"))

#Format functional terms legend
PAdown_chord <-PAdown_chord+guides(size = guide_legend("GO Terms", ncol = 1, byrow = T, 
                               override.aes = list(shape = 22, fill = unique(brewer.pal(length(PAdown_process$term),"Set2")),
                               size = 8)), fill="none")+
  theme(legend.key.size = unit(0.01, 'cm'), legend.text = element_text(size=22),
        legend.title = element_text(size=18, face="bold"))

#extracting functional terms legend
FuncLeg <- cowplot::get_legend(PAdown_chord+guides(color="none")+
     theme(legend.box.margin =unit(c(0,0,0,0), "cm"),
     legend.spacing = unit(c(-0.1,-0.1,-0.1,-0.1), "cm"),
     legend.position="right",
     legend.direction = "vertical"))

# final plot
fPA_down <- plot_grid(FCLeg,
    plot_grid(PAdown_chord+theme(legend.position = "none", 
    plot.margin = unit(c(0.1,0.1,0,0), "cm")),FuncLeg, 
    ncol=2, rel_widths = c(0.64,0.36)),
    nrow=2, rel_heights = c(0.07,0.93))


ggsave("../Enrichment/GOChord/PA_down.pdf", plot = fPA_down, dpi = 600, width = 18, height = 13)
```


# GOBubble


## Palmitic Acid

```{r PAup_bubble}

#Upregulated
PAup_bu <- GO_PA %>% 
  dplyr::filter( Regulation == "Up") %>%
  dplyr::select( ONTOLOGY, ID, Description, geneID, p.adjust ) %>% 
  dplyr::rename( category = ONTOLOGY ) %>%
  dplyr::rename( term = Description ) %>%
  dplyr::rename( genes = geneID ) %>%
  dplyr::rename( adj_pval = p.adjust ) %>%
  dplyr::mutate( genes = str_replace_all(genes, '/', ',') )

PAup.bu_circ <- circle_dat(PAup_bu, PAup_genes)

PA.bu_up <- GOBubble(PAup.bu_circ, labels = 3)
ggsave("../Enrichment/GOBubble/PA_up.pdf", plot = PA.bu_up, dpi = 600, width = 20, height = 38)
```



```{r PAdown_bubble}
#Downregulated
PAdown_bu <- GO_PA %>% 
  dplyr::filter( Regulation == "Down") %>%
  dplyr::select( ONTOLOGY, ID, Description, geneID, p.adjust ) %>% 
  dplyr::rename( category = ONTOLOGY ) %>%
  dplyr::rename( term = Description ) %>%
  dplyr::rename( genes = geneID ) %>%
  dplyr::rename( adj_pval = p.adjust ) %>%
  dplyr::mutate( genes = str_replace_all(genes, '/', ',') )

PAdown.bu_circ <- circle_dat(PAdown_bu, PAdown_genes)

PA.bu_down <- GOBubble(PAdown.bu_circ, labels = 2)
ggsave("../Enrichment/GOBubble/PA_down.pdf", plot = PA.bu_down, dpi = 600, width = 15, height = 12)
```


## Session info

```{r session_information}
Sys.time()
getwd()
  
sessionInfo()
```