---
title: "Identifying differentially expressed genes in human astrocytes exposed to Tibolone and Palmitic Acid"
output: html_notebook
author: Alexis Felipe Rojas Cruz
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.keep = 'all',
	dpi = 600,
	fig.width = 11,
	fig.height = 8.5
)
```

This notebook aims to perform differential expression (DE) analysis of human genes associated with cultures of the Normal Human Astrocyte (NHA) cell line under conditions of lipotoxicity by PA

```{r include=FALSE, comment = ""}
if (!("edgeR" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("edgeR")
}
if (!("limma" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("limma")
}
if (!("Glimma" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("Glimma")
}
if (!("BiocParallel" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("BiocParallel")
}
if (!("biomaRt" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("biomaRt")
}
if (!("RColorBrewer" %in% rownames(installed.packages()))){
    install.packages("RColorBrewer")
}
if (!("statmod" %in% rownames(installed.packages()))){
    install.packages("statmod")
}
if (!("tidyverse" %in% rownames(installed.packages()))){
    install.packages("tidyverse")
}
if (!("ggplot2" %in% rownames(installed.packages()))){
    install.packages("ggplot2")
}
if (!("ggrepel" %in% rownames(installed.packages()))){
    install.packages("ggrepel")
}
if (!("circlize" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("circlize")
}
if (!("ComplexHeatmap" %in% rownames(installed.packages()))){
    if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
    BiocManager::install("ComplexHeatmap")
}

library(edgeR)
library(limma)
library(Glimma)
library(BiocParallel)
library(biomaRt)
library(RColorBrewer)
library(statmod)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(circlize)
library(ComplexHeatmap)

```

## Data preparation

To start it requires as input the gene-level counts matrix

```{r}
DATA <- read.table(file="../Expression/astrocyte.rsem.gene_tpm.tsv", header=TRUE) 
head(DATA[,1:6])
```

We will keep the column containing "gene_id" rather than "transcript_id" for analysis and lots related to R2 and R3

```{r}
DATA <- DATA[,c(1,5:8,11:14,17:20,23:26,29:32)]
dim(DATA)
```

## Pre-filtering

Different RefSeq transcripts for the same gene symbol predominantly have the same reads. Therefore, we keep one transcript for each ID. We choose the transcript with the highest overall count

```{r}
DATA <- DATA[apply(DATA[,2:ncol(DATA)],1,max) >=2,]
sort <- order(rowSums((DATA[2:ncol(DATA)])), decreasing=TRUE)
DATA <- DATA[sort,]
DATA <- DATA[ !duplicated(DATA$gene_id), ]
nrow(DATA)
```


## Annotation

```{r gene_annotation}
mart = useMart(biomart="ENSEMBL_MART_ENSEMBL", path="/biomart/martservice", dataset="hsapiens_gene_ensembl")

ensembl = getBM(
  values = DATA$gene_id,
  filters = c("ensembl_gene_id_version"),
  attributes = c("ensembl_gene_id_version", "ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
  mart = mart
)

ensembl <- ensembl[ !duplicated(ensembl$ensembl_gene_id), ]

DATA <- ensembl %>%
  dplyr::rename( gene_id = ensembl_gene_id_version ) %>%
  dplyr::rename( symbol = hgnc_symbol ) %>%
  dplyr::rename( biotype = gene_biotype ) %>%
  merge( DATA, ., by="gene_id" ) %>%
  select( -gene_id ) %>%
  dplyr::rename( gene_id = ensembl_gene_id ) %>%
  relocate( gene_id, symbol, biotype, .before = AstroDMEM.R2A ) %>% 
  as.data.frame( . ) %>%
  unique( . )

rownames(DATA) <- DATA[,1]
dim(DATA)
```

Saving data

```{r}
write.csv(DATA, file="../Expression/Annotation/Anno_astrocyte_counts.csv", quote = F, row.names = F)
``` 


## Study design

Before starting to perform differential expression (DE) analyses, we need to prepare the design matrix related to the experimental samples. 
This dataset contains:

Treatments
-`Tibolone` (DMEM + 1.35% BSA + 2mM Carnitine + 10nM Tibolone)
-`Tibolone + Palmitic Acid` (DMEM + 1.35% BSA + 2mM Carnitine + 10nM Tibolone + 2mM Palmitic Acid)
-`Palmitic Acid` (DMEM + 1.35% BSA + 2mM Carnitine + 2mM Palmitic Acid)

Controls
-`DMEM` (DMEM)
-`VH` (DMEM + 1.35% BSA + 2mM Carnitine)

Each treatment (AstroTi, AstroPA, AstroTiPA, AstroVH, AstroDMEM) had 3 biological replicates (i.e. lots from different donors ` "R2", "R3"`) and 2 technical replicates (i.e. samples from the same lot `"A", "B"`), for a total of 30 samples. 

```{r}
condition <- do.call(rbind,strsplit(colnames(DATA)[4:ncol(DATA)], split="[.]"))[,1]
replicates <- do.call(rbind,strsplit(colnames(DATA)[4:ncol(DATA)], split="[.]"))[,2]
colData <- data.frame(condition=factor(condition, levels=c("AstroDMEM", "AstroVH", "AstroTi", "AstroTiPA", "AstroPA")),
                   replicates=factor(replicates, levels=c("R2A", "R2B", "R3A", "R3B")))
colData$group <- factor(paste(condition, replicates, sep="_"))
rownames(colData) <- colnames(DATA)[4:ncol(DATA)]
print(colData)
```

## Data integration

We then use this vector and the gene counts to create a DGEList, which is the object that edgeR uses for storing the data from a differential expression experiment.

```{r}
countsMat <- round(as.matrix(DATA[4:ncol(DATA)]))
rownames(countsMat) <- DATA[,1]
ddsNHA <- DGEList(counts = countsMat,
                  group = colData$condition)
ddsNHA$genes <- data.frame(DATA[1:3])
ddsNHA$samples
```


```{r}
cols <- as.numeric(ddsNHA$samples$group)+2
barplot(ddsNHA$samples$lib.size*1e-04, col = cols, names = colnames(ddsNHA),
        las = 2, ann = FALSE, cex.names = 0.7, cex.axis = 0.8)
mtext(side = 1, text = "Samples", line = 4)
mtext(side = 2, text = "Library size (millions)", line = 2)
title("Barplot of library sizes")
```

## Filtering

We should also remove genes that are unexpressed or very lowly expressed in the samples. Genes with a minimum requirement of 20 count per million (CPM) across libraries for each group have been kept

```{r}
keep <- filterByExpr(ddsNHA, min.count = 20)
table(keep)
ddsNHA <- ddsNHA[keep, , keep.lib.sizes=FALSE]
```


## Normalization

The TMM normalization is applied to account for the compositional biases

```{r}
ddsNHA <- calcNormFactors(ddsNHA)
ddsNHA$samples
```


## Quality control

Next we’ll use boxplots to check the distribution of the read counts on the log2 scale. We can use the cpm function to get log2 counts per million, which are corrected for the different library sizes

```{r warning=FALSE}
par(mfrow=c(1,1))
logcounts <- cpm(ddsNHA,log=TRUE)
nsamples <- ncol(logcounts)
col <- brewer.pal(nsamples, "Paired")
boxplot(logcounts, las=2, col=col, main="", cex.axis = 0.75, cex = 0.7)
abline(h=median(logcounts),col="blue")
title(main="Normalized data",ylab="Log-cpm")
```

## Data exploration

Before proceeding with the computations for differential expression, it is possible to produce a plot showing the sample relations based on multidimensional scaling (PCA). This is something that we will cover in much more detail in a later lecture

```{r}
points <- c(0,1,2,15,8)
colors <- c("blue", "darkgreen", "red", "yellow", "violet")
plotMDS(ddsNHA, col = colors[colData$condition], pch = points[colData$condition])
legend("bottomright", legend = levels(colData$condition), pch=points, col = colors, ncol = 2, cex = 0.9)

```

## The design matrix

```{r}
group <- colData$condition
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
design
```

## Estimating the dispersion

DE analysis using the NB model, which estimates the dispersion parameter for each gene. Estimating the common dispersion gives an idea of overall variability across the genome for this dataset.

```{r}
ddsNHA <- estimateDisp(ddsNHA, design, robust=TRUE)
plotBCV(ddsNHA)
```

For the QL dispersions, estimation can be performed using the `glmQLFit` function

```{r}
fit <- glmQLFit(ddsNHA, design, robust=TRUE)
head(fit$coefficients)
plotQLDisp(fit)
```


## Differential expression

Now we will find genes differentially expressed among multiple combinations


## Contrasts

For this study we will perform 12 comparisons including:


- `AstroVH vs AstroDMEM`
- `AstroDMEM vs AstroVH`
- `AstroTi vs AstroDMEM`
- `AstroTi vs AstroVH`
- `(AstroTi - AstroDMEM) vs (AstroVH - AstroDMEM)`
- `AstroPA vs AstroDMEM`
- `AstroPA vs AstroVH`
- `AstroTiPA vs AstroDMEM`
- `AstroTiPA vs AstroVH`
- `AstroTiPA vs AstroPA`
- `AstroPA vs AstroTi`
- `(AstroPA - AstroDMEM) vs (AstroTi - AstroDMEM)`
- `(AstroPA - AstroVH) vs (AstroTi - AstroVH)`
- `(AstroPA - AstroDMEM) vs (AstroTiPA - AstroDMEM)`
- `(AstroPA - AstroVH) vs (AstroTiPA - AstroVH)`

The design of my contrasts is as follows

```{r}
my.contrasts <- makeContrasts(
    # Vehicle vs DMEM
    VHvsDMEM = AstroVH-AstroDMEM,
    # DMEM vs Vehicle
    DMEMvsVH = AstroDMEM-AstroVH,
    # Tibolone vs DMEM
    TivsDMEM = AstroTi-AstroDMEM,
    # Tibolone vs VH
    TivsVH = AstroTi-AstroVH,
    # Difference between Tibolone vs DMEM and Vehicle vs DMEM
    Ti_DMEMvsVHDMEM = (AstroTi-AstroDMEM)-(AstroVH-AstroDMEM),
    # Palmitic Acid vs DMEM
    PAvsDMEM = AstroPA-AstroDMEM,
    # Palmitic Acid vs VH
    PAvsVH = AstroPA-AstroVH,
    # Tibolone + Palmitic Acid vs DMEM
    TiPAvsDMEM = AstroTiPA-AstroDMEM,
    # Tibolone + Palmitic Acid vs VH
    TiPAvsVH = AstroTiPA-AstroVH,
    # Tibolone + Palmitic Acid vs Palmitic Acid
    TiPAvsPA = AstroTiPA-AstroPA,
    # Palmitic Acid vs Tibolone
    PAvsTi = AstroPA-AstroTi,
    # Difference between Palmitic Acid vs DMEM and Tibolone vs DMEM
    PA_TivsDMEM = (AstroPA-AstroDMEM)-(AstroTi-AstroDMEM),
    # Difference between Palmitic Acid vs VH and Tibolone vs VH
    PA_TivsVH = (AstroPA-AstroVH)-(AstroTi-AstroVH),
    # Difference between Palmitic Acid vs DMEM and Tibolone + Palmitic Acid vs DMEM
    PA_TiPAvsDMEM = (AstroPA-AstroDMEM)-(AstroTiPA-AstroDMEM),
    # Difference between Palmitic Acid vs VH and Tibolone + Palmitic Acid vs VH
    PA_TiPAvsVH = (AstroPA-AstroVH)-(AstroTiPA-AstroVH),   
    levels=design)
```

## Functions

```{r}
# pretty format for edgeR results
format_edgeR_results <- function( edgeR_results ) {
  as.data.frame( edgeR_results ) %>%
    filter( !is.na( PValue ) ) %>%
    filter( !is.na( FDR ) ) %>%
    select( gene_id, symbol, biotype, logFC, logCPM, F, PValue, FDR ) %>%
    arrange( PValue, FDR, gene_id ) %>%
    return( . )
}
```

```{r}
# Report differential expression

summarize_pretty_edg <- function( pretty_edg, lfc_cutoff ) {
  down <- pretty_edg %>%
    filter( FDR <= 0.05 & logFC <= 1.0 ) %>%
    pull( gene_id ) %>%
    length( . )

  down_fc <- pretty_edg %>%
    filter( FDR <= 0.05 & logFC <= ( -1 * lfc_cutoff ) ) %>%
    pull( gene_id ) %>%
    length( . )

  up <- pretty_edg %>%
    filter( FDR <= 0.05 & logFC >= 1.0 ) %>%
    pull( gene_id ) %>%
    length( . )

  up_fc <- pretty_edg %>%
    filter( FDR <= 0.05 & logFC >= lfc_cutoff ) %>%
    pull( gene_id ) %>%
    length( . )

data.frame( Change = c( 'Up', paste0( 'Up - min ', sprintf( fmt="%.2f", lfc_cutoff ) ), 'Down', paste0( 'Down - min ', sprintf( fmt="%.2f", lfc_cutoff ) ) ), n = c( up, up_fc, down, down_fc ) ) %>%
    return( . )
}
```

```{r}

# ggplot modifiers

gg_bigger_texts = theme(
  axis.title = element_text( size = 20 ),
  axis.text = element_text( size = 20 ),
  legend.text = element_text( size = 14 ),
  legend.title = element_text( size = 15 ),
  plot.title = element_text( size = 20 ),
  strip.text.x = element_text( size = 17, margin = margin( b = 5, t = 5 ) ),
  strip.text.y = element_text( size = 15 )
)

gg_no_legend = theme(
  legend.position='none'
)

gg_no_grid = theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
)

gg_center_title = theme(
  plot.title = element_text( hjust = 0.5 )
)

# volcano plots 

make_ggplot_volcano <- function( deg_df, case_name, control_name, axis_steps = 2, lfc_cutoff = 1, FDR_cutoff = 0.05, max_label = 20 )
{

  # set significance threshold 

  deg_df <- deg_df %>%
    mutate( Significant = case_when(
      FDR < FDR_cutoff & abs( logFC ) >= lfc_cutoff ~ "Large",
      FDR < FDR_cutoff ~ "Modest",
      TRUE ~ "Not" ) ) %>%
    mutate( Significant = factor( Significant, levels=c( "Not", "Modest", "Large" ) ) )


  # set values for square x axis

  x_volcano_value <- ( abs( All_VHvsDMEM$logFC[ is.finite( All_VHvsDMEM$logFC ) ] ) + 0.05 ) %>%
    max( . ) %>%
    round( ., 1 )

  if ( x_volcano_value < 1.0 ) {
    x_volcano_value = 1.0
  }
  
  x_num_for_limits <- round( x_volcano_value, 0 )

  x_volcano_low <- x_volcano_value * -1
  x_volcano_high <- x_volcano_value

  x_break_list <- seq( -1 * x_num_for_limits, x_num_for_limits, by = axis_steps )

 
  # plot lines

  horizontal_line <- log10( FDR_cutoff ) * -1
  vertical_line_1 <- lfc_cutoff
  vertical_line_2 <- vertical_line_1 * -1


  # actually make the volcano plots

  plot_volcano <- ggplot( deg_df, aes( x=logFC, y=-log10( FDR ), colour=Significant ) ) +
    scale_colour_manual( values = c( "darkgray", "mediumseagreen", "red" ) ) +
    scale_x_continuous( limits = c( x_volcano_low, x_volcano_high ), breaks = x_break_list ) +
    theme_bw() +
    gg_bigger_texts +
    gg_no_legend +
    gg_no_grid +
    gg_center_title +
    geom_point( size=1.2 ) +
    geom_hline( yintercept = horizontal_line, linetype=2 ) +
    geom_vline( xintercept=c( vertical_line_1, vertical_line_2 ), linetype=2 ) +
    geom_text_repel( data=subset( deg_df, Significant == "Large" )[c(1:max_label),],
                     colour="black", aes( label=symbol ), size=3) +
    xlab( parse( text=paste0( "log[2]~(", case_name, "/", control_name, ")" ) ) ) +
    ylab( parse( text = paste0( "-log[10]~(Adj.~p-value)" ) ) )

  return( plot_volcano )
}
```


```{r}
# Creating a list of DEGs

getting_DEGs <- function( deg_df, FDR_cutoff, lfc_cutoff ) {
 deg_df <- deg_df %>% 
    mutate( regulation = case_when(
      FDR < FDR_cutoff & logFC >= lfc_cutoff ~ "Up",
      FDR < FDR_cutoff & logFC <= ( -1 * lfc_cutoff ) ~ "Down",
      TRUE ~ "Not" ) ) %>%
   filter( regulation != "Not" ) %>% 
   arrange( desc(logFC) ) %>% 
   return( . )
}
```

## Vehicule vs DMEM

Our first comparison aims to identify differentially expressed genes (DEGs) for both controls: Vehicule versus DMEM

```{r comparison VHvsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"VHvsDMEM"])
VHvsDMEM <- topTags(res, n = nrow(fit$counts))
# adjust p-values FDR
VHvsDMEM$table$FDR <- p.adjust(VHvsDMEM$table$PValue, method = "BH")
# showing results
topTags(res)
summary(decideTests(res))
```

Now we will format our results in a fancier layout

```{r format VHvsDMEM}
All_VHvsDMEM <- format_edgeR_results(VHvsDMEM)
head(All_VHvsDMEM)
```

DEGs are selected by keeping those having an absolute log2 fold change at 1 and a FDR lower than 0.05.

```{r report of DEGs VHvsDMEM}
deg_df_VHvsDMEM <- summarize_pretty_edg(All_VHvsDMEM, 1)
head(deg_df_VHvsDMEM)
```

```{r volcano VHvsDMEM, message=FALSE, warning=FALSE}
volcano_VHvsDMEM <- make_ggplot_volcano( All_VHvsDMEM, case_name = "VH", control_name = "DMEM" )
volcano_VHvsDMEM
ggsave("../Expression/Volcano/Vol_VHvsDMEM.pdf", plot = volcano_VHvsDMEM)
```


```{r save VHvsDMEM}
VHvsDMEMfilt <- getting_DEGs( All_VHvsDMEM, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(VHvsDMEMfilt)
write.csv(VHvsDMEMfilt, file="../Expression/DEGs/VHvsDMEM.csv", quote = F, row.names = F)
```

## DMEM vs VH


```{r comparison DMEMvsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"DMEMvsVH"])
DMEMvsVH <- topTags(res, n = nrow(fit$counts))
DMEMvsVH$table$FDR <- p.adjust(DMEMvsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```


```{r format DMEMvsVH}
All_DMEMvsVH <- format_edgeR_results(DMEMvsVH)
head(All_DMEMvsVH)
```


```{r report of DEGs DMEMvsVH}
deg_df_DMEMvsVH <- summarize_pretty_edg(All_DMEMvsVH, 1)
head(deg_df_DMEMvsVH)
```

```{r volcano DMEMvsVH, message=FALSE, warning=FALSE}
volcano_DMEMvsVH <- make_ggplot_volcano( All_DMEMvsVH, case_name = "DMEM", control_name = "VH" )
volcano_DMEMvsVH
ggsave("../Expression/Volcano/Vol_DMEMvsVH.pdf", plot = volcano_DMEMvsVH)
```

```{r save DMEMvsVH}
DMEMvsVHfilt <- getting_DEGs( All_DMEMvsVH, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(DMEMvsVHfilt)
write.csv(DMEMvsVHfilt, file="../Expression/DEGs/DMEMvsVH.csv", quote = F, row.names = F)
```

## `Tibolone vs DMEM`


```{r comparison TivsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"TivsDMEM"])
TivsDMEM <- topTags(res, n = nrow(fit$counts))
TivsDMEM$table$FDR <- p.adjust(TivsDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format TivsDMEM}
All_TivsDMEM <- format_edgeR_results(TivsDMEM)
head(All_TivsDMEM)
```


```{r report of DEGs TivsDMEM}
deg_df_TivsDMEM <- summarize_pretty_edg(All_TivsDMEM, 1)
head(deg_df_TivsDMEM)
```

```{r volcano TivsDMEM, message=FALSE, warning=FALSE}
volcano_TivsDMEM <- make_ggplot_volcano( All_TivsDMEM, case_name = "TIB", control_name = "DMEM" )
volcano_TivsDMEM
ggsave("../Expression/Volcano/Vol_TivsDMEM.pdf", plot = volcano_TivsDMEM)
```
As a result, for this comparison we found: `Up` = 0 DEGs and `Down` = 0 DEGs

## Tibolone vs VH

```{r comparison TivsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"TivsVH"])
TivsVH <- topTags(res, n = nrow(fit$counts))
TivsVH$table$FDR <- p.adjust(TivsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```


```{r format TivsVH}
All_TivsVH <- format_edgeR_results(TivsVH)
head(All_TivsVH)
```


```{r report of DEGs TivsVH}
deg_df_TivsVH <- summarize_pretty_edg(All_TivsVH, 1)
head(deg_df_TivsVH)
```

Volcano plot for TivsVH

```{r volcano TivsVH, message=FALSE, warning=FALSE}
volcano_TivsVH <- make_ggplot_volcano( All_TivsVH, case_name = "TIB", control_name = "VH" )
volcano_TivsVH
ggsave("../Expression/Volcano/Vol_TIBvsVH.pdf", plot = volcano_TivsVH)
```


```{r save TivsVH}
TivsVHfilt <- getting_DEGs( All_TivsVH, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(TivsVHfilt)
write.csv(TivsVHfilt, file="../Expression/DEGs/TIBvsVH.csv", quote = F, row.names = F)
```


## Difference between Tibolone vs DMEM and VH vs DMEM

```{r comparison Ti_DMEMvsVHDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"Ti_DMEMvsVHDMEM"])
Ti_DMEMvsVHDMEM <- topTags(res, n = nrow(fit$counts))
Ti_DMEMvsVHDMEM$table$FDR <- p.adjust(Ti_DMEMvsVHDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format Ti_DMEMvsVHDMEM}
All_Ti_DMEMvsVHDMEM <- format_edgeR_results(Ti_DMEMvsVHDMEM)
head(All_Ti_DMEMvsVHDMEM)
```

```{r report of DEGs Ti_DMEMvsVHDMEM}
deg_df_Ti_DMEMvsVHDMEM <- summarize_pretty_edg(All_Ti_DMEMvsVHDMEM, 1)
head(deg_df_Ti_DMEMvsVHDMEM)
```

```{r volcano Ti_DMEMvsVHDMEM, message=FALSE, warning=FALSE}
volcano_Ti_DMEMvsVHDMEM <- make_ggplot_volcano( All_Ti_DMEMvsVHDMEM, case_name = "TIB-DMEM", control_name = "TIB-VH-DMEM" )
volcano_Ti_DMEMvsVHDMEM
ggsave("../Expression/Volcano/Vol_TIB_DMEMvsVHDMEM.pdf", plot = volcano_Ti_DMEMvsVHDMEM)
```

```{r save Ti_DMEMvsVHDMEM}
Ti_DMEMvsVHDMEMfilt <- getting_DEGs( All_Ti_DMEMvsVHDMEM, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(Ti_DMEMvsVHDMEMfilt)
write.csv(Ti_DMEMvsVHDMEMfilt, file="../Expression/DEGs/TIB_DMEMvsVHDMEM.csv", quote = F, row.names = F)
```



## Palmitic Acid vs DMEM

```{r comparison PAvsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PAvsDMEM"])
PAvsDMEM <- topTags(res, n = nrow(fit$counts))
PAvsDMEM$table$FDR <- p.adjust(PAvsDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PAvsDMEM}
All_PAvsDMEM <- format_edgeR_results(PAvsDMEM)
head(All_PAvsDMEM)
```

```{r report of DEGs PAvsDMEM}
deg_df_PAvsDMEM <- summarize_pretty_edg(All_PAvsDMEM, 1)
head(deg_df_PAvsDMEM)
```

```{r volcano PAvsDMEM, message=FALSE, warning=FALSE}
volcano_PAvsDMEM <- make_ggplot_volcano( All_PAvsDMEM, case_name = "PA", control_name = "DMEM" )
volcano_PAvsDMEM
ggsave("../Expression/Volcano/Vol_PAvsDMEM.pdf", plot = volcano_PAvsDMEM)
```

```{r save PAvsDMEM}
PAvsDMEMfilt <- getting_DEGs( All_PAvsDMEM, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(PAvsDMEMfilt)
write.csv(PAvsDMEMfilt, file="../Expression/DEGs/PAvsDMEM.csv", quote = F, row.names = F)
```


## Palmitic Acid vs VH

```{r comparison PAvsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PAvsVH"])
PAvsVH <- topTags(res, n = nrow(fit$counts))
PAvsVH$table$FDR <- p.adjust(PAvsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PAvsVH}
All_PAvsVH <- format_edgeR_results(PAvsVH)
head(All_PAvsVH)
```

```{r report of DEGs PAvsVH}
deg_df_PAvsVH <- summarize_pretty_edg(All_PAvsVH, 1)
head(deg_df_PAvsVH)
```

```{r volcano PAvsVH, message=FALSE, warning=FALSE}
volcano_PAvsVH <- make_ggplot_volcano( All_PAvsVH, case_name = "PA", control_name = "VH" )
volcano_PAvsVH
ggsave("../Expression/Volcano/Vol_PAvsVH.pdf", plot = volcano_PAvsVH)
```

```{r save PAvsVH}
PAvsVHfilt <- getting_DEGs( All_PAvsVH, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(PAvsVHfilt)
write.csv(PAvsVHfilt, file="../Expression/DEGs/PAvsVH.csv", quote = F, row.names = F)
```


## Tibolone + Palmitic Acid vs DMEM

```{r comparison TiPAvsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"TiPAvsDMEM"])
TiPAvsDMEM <- topTags(res, n = nrow(fit$counts))
TiPAvsDMEM$table$FDR <- p.adjust(TiPAvsDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format TiPAvsDMEM}
All_TiPAvsDMEM <- format_edgeR_results(TiPAvsDMEM)
head(All_TiPAvsDMEM)
```

```{r report of DEGs TiPAvsDMEM}
deg_df_TiPAvsDMEM <- summarize_pretty_edg(All_TiPAvsDMEM, 1)
head(deg_df_TiPAvsDMEM)
```

```{r volcano TiPAvsDMEM, message=FALSE, warning=FALSE}
volcano_TiPAvsDMEM <- make_ggplot_volcano( All_TiPAvsDMEM, case_name = "TIBPA", control_name = "DMEM" )
volcano_TiPAvsDMEM
ggsave("../Expression/Volcano/Vol_TIBPAvsDMEM.pdf", plot = volcano_TiPAvsDMEM)
```

```{r save TiPAvsDMEM}
TiPAvsDMEMfilt <- getting_DEGs( All_TiPAvsDMEM, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(TiPAvsDMEMfilt)
write.csv(TiPAvsDMEMfilt, file="../Expression/DEGs/TIBPAvsDMEM.csv", quote = F, row.names = F)
```


## Tibolone + Palmitic Acid vs VH

```{r comparison TiPAvsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"TiPAvsVH"])
TiPAvsVH <- topTags(res, n = nrow(fit$counts))
TiPAvsVH$table$FDR <- p.adjust(TiPAvsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format TiPAvsVH}
All_TiPAvsVH <- format_edgeR_results(TiPAvsVH)
head(All_TiPAvsVH)
```

```{r report of DEGs TiPAvsVH}
deg_df_TiPAvsVH <- summarize_pretty_edg(All_TiPAvsVH, 1)
head(deg_df_TiPAvsVH)
```

```{r volcano TiPAvsVH, message=FALSE, warning=FALSE}
volcano_TiPAvsVH <- make_ggplot_volcano( All_TiPAvsVH, case_name = "TIPA", control_name = "VH" )
volcano_TiPAvsVH
ggsave("../Expression/Volcano/Vol_TIBPAvsVH.pdf", plot = volcano_TiPAvsVH)
```

```{r save TiPAvsVH}
TiPAvsVHfilt <- getting_DEGs( All_TiPAvsVH, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(TiPAvsVHfilt)
write.csv(TiPAvsVHfilt, file="../Expression/DEGs/TIBPAvsVH.csv", quote = F, row.names = F)
```


## Palmitic Acid vs Tibolone

```{r comparison PAvsTi}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PAvsTi"])
PAvsTi <- topTags(res, n = nrow(fit$counts))
PAvsTi$table$FDR <- p.adjust(PAvsTi$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PAvsTi}
All_PAvsTi <- format_edgeR_results(PAvsTi)
head(All_PAvsTi)
```

```{r report of DEGs PAvsTi}
deg_df_PAvsTi <- summarize_pretty_edg(All_PAvsTi, 1)
head(deg_df_PAvsTi)
```

```{r volcano PAvsTi, message=FALSE, warning=FALSE}
volcano_PAvsTi <- make_ggplot_volcano( All_PAvsTi, case_name = "PA", control_name = "TIB" )
volcano_PAvsTi
ggsave("../Expression/Volcano/Vol_PAvsTIB.pdf", plot = volcano_PAvsTi)
```

```{r save PAvsTi}
PAvsTifilt <- getting_DEGs( All_PAvsTi, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(PAvsTifilt)
write.csv(PAvsTifilt, file="../Expression/DEGs/PAvsTIB.csv", quote = F, row.names = F)
```


## `Tibolone + Palmitic Acid vs Palmitic Acid`

```{r comparison TiPAvsPA}
res <- glmQLFTest(fit, contrast=my.contrasts[,"TiPAvsPA"])
TiPAvsPA <- topTags(res, n = nrow(fit$counts))
TiPAvsPA$table$FDR <- p.adjust(TiPAvsPA$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format TiPAvsPA}
All_TiPAvsPA <- format_edgeR_results(TiPAvsPA)
head(All_TiPAvsPA)
```

```{r report of DEGs TiPAvsPA}
deg_df_TiPAvsPA <- summarize_pretty_edg(All_TiPAvsPA, 1)
head(deg_df_TiPAvsPA)
```

```{r volcano TiPAvsPA, message=FALSE, warning=FALSE}
volcano_TiPAvsPA <- make_ggplot_volcano( All_TiPAvsPA, case_name = "TIBPA", control_name = "PA" )
volcano_TiPAvsPA
ggsave("../Expression/Volcano/Vol_TIBPAvsPA.pdf", plot = volcano_TiPAvsPA)
```

For this comparison we found: `Up` = 0 DEGs and `Down` = 0 DEGs


## Difference between Palmitic Acid vs DMEM and Tibolone vs DMEM

```{r comparison PA_TivsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PA_TivsDMEM"])
PA_TivsDMEM <- topTags(res, n = nrow(fit$counts))
PA_TivsDMEM$table$FDR <- p.adjust(PA_TivsDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PA_TivsDMEM}
All_PA_TivsDMEM <- format_edgeR_results(PA_TivsDMEM)
head(All_PA_TivsDMEM)
```

```{r report of DEGs PA_TivsDMEM}
deg_df_PA_TivsDMEM <- summarize_pretty_edg(All_PA_TivsDMEM, 1)
head(deg_df_PA_TivsDMEM)
```

```{r volcano PA_TivsDMEM, message=FALSE, warning=FALSE}
volcano_PA_TivsDMEM <- make_ggplot_volcano( All_PA_TivsDMEM, case_name = "PA-DMEM", control_name = "TIB-DMEM" )
volcano_PA_TivsDMEM
ggsave("../Expression/Volcano/Vol_PA_TIBvsDMEM.pdf", plot = volcano_PA_TivsDMEM)
```

```{r save PA_TivsDMEM}
PA_TivsDMEMfilt <- getting_DEGs( All_PA_TivsDMEM, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(PA_TivsDMEMfilt)
write.csv(PA_TivsDMEMfilt, file="../Expression/DEGs/PA_TIBvsDMEM.csv", quote = F, row.names = F)
```


## Difference between Palmitic Acid vs VH and Tibolone vs VH

```{r comparison PA_TivsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PA_TivsVH"])
PA_TivsVH <- topTags(res, n = nrow(fit$counts))
PA_TivsVH$table$FDR <- p.adjust(PA_TivsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PA_TivsVH}
All_PA_TivsVH <- format_edgeR_results(PA_TivsVH)
head(All_PA_TivsVH)
```

```{r report of DEGs PA_TivsVH}
deg_df_PA_TivsVH <- summarize_pretty_edg(All_PA_TivsVH, 1)
head(deg_df_PA_TivsVH)
```

```{r volcano PA_TivsVH, message=FALSE, warning=FALSE}
volcano_PA_TivsVH <- make_ggplot_volcano( All_PA_TivsVH, case_name = "PA-VH", control_name = "TIB-VH" )
volcano_PA_TivsVH
ggsave("../Expression/Volcano/Vol_PA_TIBvsVH.pdf", plot = volcano_PA_TivsVH)
```

```{r save PA_TivsVH}
PA_TivsVHfilt <- getting_DEGs( All_PA_TivsVH, FDR_cutoff = 0.05, lfc_cutoff = 1 )
head(PA_TivsVHfilt)
write.csv(PA_TivsVHfilt, file="../Expression/DEGs/PA_TIBvsVH.csv", quote = F, row.names = F)
```


## `Difference between Palmitic Acid vs DMEM and Tibolone + Palmitic Acid vs DMEM`

```{r comparison PA_TiPAvsDMEM}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PA_TiPAvsDMEM"])
PA_TiPAvsDMEM<- topTags(res, n = nrow(fit$counts))
PA_TiPAvsDMEM$table$FDR <- p.adjust(PA_TiPAvsDMEM$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PA_TiPAvsDMEM}
All_PA_TiPAvsDMEM <- format_edgeR_results(PA_TiPAvsDMEM)
head(All_PA_TiPAvsDMEM)
```

```{r report of DEGs PA_TiPAvsDMEM}
deg_df_PA_TiPAvsDMEM <- summarize_pretty_edg(All_PA_TiPAvsDMEM, 1)
head(deg_df_PA_TiPAvsDMEM)
```

For this comparison we found: `Up` = 0 DEGs and `Down` = 0 DEGs


## `Difference between Palmitic Acid vs VH and Tibolone + Palmitic Acid vs VH`

```{r comparison PA_TiPAvsVH}
res <- glmQLFTest(fit, contrast=my.contrasts[,"PA_TiPAvsVH"])
PA_TiPAvsVH <- topTags(res, n = nrow(fit$counts))
PA_TiPAvsVH$table$FDR <- p.adjust(PA_TiPAvsVH$table$PValue, method = "BH")

topTags(res)
summary(decideTests(res))
```

```{r format PA_TiPAvsVH}
All_PA_TiPAvsVH <- format_edgeR_results(PA_TiPAvsVH)
head(All_PA_TiPAvsVH)
```

```{r report of DEGs PA_TiPAvsVH}
deg_df_PA_TiPAvsVH <- summarize_pretty_edg(All_PA_TiPAvsVH, 1)
head(deg_df_PA_TiPAvsVH)
```

For this comparison we found: `Up` = 0 DEGs and `Down` = 0 DEGs


## Heatmap

```{r heatmap, message=FALSE, warning=FALSE}
logCPM <- cpm(ddsNHA, prior.count=2, log=TRUE)

topDEGs <- c(rownames(TiPAvsVHfilt),
             rownames(PAvsVHfilt))

logCPM <- logCPM[,c(1:16)]
col_order <- c(13:16,5:8,9:12,1:4)
logCPM <- logCPM[, col_order]
colnames(logCPM) <- factor(c("TiPA-R1 ", "TiPA-R2", "TiPA-R3", "TiPA-R4",
                             "PA-R1 ", "PA-R2", "PA-R3", "PA-R4",
                             "Ti-R1", "Ti-R2", "Ti-R3", "Ti-R4",
                             "DMEM-R1", "DMEM-R2", "DMEM-R3", "DMEM-R4"))

logCPM <- logCPM[rownames(logCPM)%in%topDEGs,]
logCPM <- t(scale(t(logCPM)))

# get log2change and p-values

logFC <- All_TiPAvsVH %>% 
  dplyr::select(TiPA = logFC, gene_id) %>%
  merge(All_PAvsVH, ., by = "gene_id") %>%
  dplyr::select(TiPA, PA = logFC, gene_id) %>%
  tibble::column_to_rownames(var = "gene_id")


#select only significant logFC

logFC <- logFC[rownames(logFC)%in%topDEGs,]


# ordering

logCPM <- logCPM[order(logFC$TiPA, decreasing = T), ]
logFC <- logFC[order(logFC$TiPA, decreasing = T), ]
```


```{r}
orden <- c(rep("TiPA", 4), rep("PA", 4), rep("Ti", 4), rep("DMEM", 4))
level <- c("TiPA", "PA", "Ti", "DMEM")
c1 =  circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
c2 =  circlize::colorRamp2(c(-2, 0, 2), c("darkblue", "white", "green"))
ht1 <- Heatmap(logCPM, name = "Log CPM",
               show_row_names = F, show_column_names = F,
               col=c1, cluster_rows = F,
               show_heatmap_legend = F, cluster_columns = F,
               row_title_side = "right", row_title_rot = 0,
               border = TRUE, 
               column_names_max_height=max_text_width(rownames(logCPM)),
               column_split = factor(orden, levels = level), 
               row_gap = unit(0, "cm"))


ht2 <- Heatmap(logFC, name = "Log2 FC",
               col = c2,
               cluster_rows = F, cluster_columns = F, show_heatmap_legend = F,
               show_row_names = F, border = TRUE, row_title_side = "right", row_title_rot = 0,
               column_split = factor(c("TiPA", "PA"), levels = level),
               show_column_names = F, width = unit(6, "cm"))

ht_list = ht1 + ht2 
ht_list

lgd1 = Legend(col_fun = c1, title = expression("Log CPM"), labels_gp = gpar(fontsize = 20),
             title_gp = gpar(fontsize = 24),
             at = c(-3, 0, 3), legend_width = unit(4, "cm"), grid_width =  unit(0.8, "cm"),
             labels = c("-3", "  0", "  3"), legend_height = unit(5, "cm"),
             title_position = "leftcenter-rot")
lgd2 = Legend(col_fun = c2, title = expression("Log"[2]*" FC"), labels_gp = gpar(fontsize = 20),
             title_gp = gpar(fontsize = 24),grid_width =  unit(0.8, "cm"),
             at = c(-8, 0, 8), legend_width = unit(4, "cm"),
             labels = c("-8", "  0", "  8"), legend_height = unit(5, "cm"),
             title_position = "leftcenter-rot")


pdf("../Expression//Heatmap/heatmap.pdf", width = 10, height = 20)
draw(ht_list, ht_gap = unit(1, "cm"), padding = unit(c(0.2, 0.2, 0.2, 5.5), "cm"))
draw(lgd1, x = unit(22, "cm"), y = unit(33.35, "cm"), just = c("left", "top"))
draw(lgd2, x = unit(22, "cm"), y = unit(23.3, "cm"), just = c("left", "top"))
dev.off()

```

## Getting expressed genes

In order to perform an overrepresentation analysis (ORA), we will retrieve all genes were selected from already had unreliably detected, low count genes removed

```{r aal_genes}

allgenesfilt <- ddsNHA[["genes"]][["gene_id"]] %>%
  data.frame( . ) %>% 
  dplyr::rename( ., gene_id = . )

write.csv(allgenesfilt, file="../Expression/Annotation/allgenesfilt.csv", quote = F, row.names = F)
```

## Session info

```{r session_information}
Sys.time()
getwd()
  
sessionInfo()
```