---
title: "Identification of overlapping differentially expressed genes associated with Palmitic Acid"
output: html_notebook
author: Alexis Felipe Rojas Cruz
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.keep = 'all',
	dpi = 600,
	fig.width = 11,
	fig.height = 8.5
)
```

After expression analysis, we obtained the following results

|No.|Versus	                                                  |Total DEGs	  |Up DEGs     |Down DEGs    |
|:--|:--------------------------------------------------------|:-----------:|:----------:|:-----------:|
|1	|AstroVH vs AstroDMEM                                     |148	        |101         |47           |
|2	|AstroDMEM vs AstroVH                                     |148	        |47          |101          |
|3	|AstroTi vs AstroDMEM                                     |0	          |0	         |0            |
|4	|AstroTi vs AstroVH 	                                    |148          |53	         |95           |
|5	|(AstroTi vs AstroDMEM) vs (AstroVH vs AstroDMEM) 	      |148          |53	         |95           |
|6	|AstroPA vs AstroDMEM                                   	|1064	        |538	       |526          |
|7  |AstroPA vs AstroVH	                                      |739	        |366	       |373          |
|8	|AstroTiPA vs AstroDMEM                    	              |1302	        |590	       |712          |
|9  |AstroTiPA vs AstroVH	                                    |1052	        |442	       |610          |
|10	|AstroPA vs AstroTi	                                      |1128	        |527  	     |601          |
|11	|AstroPA vs AstroTiPA	                                    |0	          |0	         |0            |
|12	|(AstroPA - AstroDMEM) vs (AstroTi - AstroDMEM)	          |1128	        |527  	     |601          |
|13	|(AstroPA - AstroVH) vs (AstroTi - AstroVH)             	|1128	        |527  	     |601          |
|14	|(AstroPA - AstroDMEM) vs (AstroTiPA - AstroDMEM)	        |0	          |0	         |0            |
|15	|(AstroPA - AstroVH) vs (AstroTiPA - AstroVH)	            |0	          |0	         |0            |


```{r include=FALSE, comment = ""}
if (!("ggVennDiagram" %in% rownames(installed.packages()))){
    install.packages("ggVennDiagram")
}
if (!("sf" %in% rownames(installed.packages()))){
    install.packages("sf")
}
if (!("RColorBrewer" %in% rownames(installed.packages()))){
    install.packages("RColorBrewer")
}
if (!("tidyverse" %in% rownames(installed.packages()))){
    install.packages("tidyverse")
}
if (!("ggplot2" %in% rownames(installed.packages()))){
    install.packages("ggplot2")
}
if (!("ggrepel" %in% rownames(installed.packages()))){
    install.packages("ggrepel")
}
if (!("devtools" %in% rownames(installed.packages()))){
    install.packages("devtools")
}
if (!("ggsci" %in% rownames(installed.packages()))){
    install.packages("ggsci")
}

library(ggVennDiagram)
library(sf)
library(RColorBrewer)
library(tidyverse)
library(ggplot2)
library(devtools)
library(ggsci)
```

## Loading data

Here we will analyze differentially expressed tibolone genes which have a neuroprotective effect against palmitic acid-treated astrocytes, considered as one of the most abundant acids in our diet

`Tibolone` = Considering that VH may regulate DEGs that do not correspond to the action of Tibolone, it was compared with DMEMvsVH to deduce possible DEGs induced by Tibolone.

`Tibolone + Palmitic Acid` = To verify if in effect the DEGs resulting from the TivsVH contrast are associated with Tibolone, we compared them with the unique Tibolone contrast which is TiPAvsVH to assess whether the DEGs match or not. 

`Palmitic Acid` = The lipotoxic effect of Palmitic Acid on astrocytes will be evaluated

Thus, the following contrast be analyzed:

`TIBOLONE`
- `AstroVH vs AstroDMEM`
- `AstroTi vs AstroVH`

`TIBOLONE + PALMITIC ACID`
- `AstroTiPA vs AstroVH`
- `AstroPA vs AstroVH`
- `AstroVH vs AstroDMEM`
- `AstroTi vs AstroVH`

`PALMITIC ACID`
- `AstroTiPA vs AstroVH`
- `AstroPA vs AstroVH`


```{r data}
TivsVH <- read.csv("../Expression/DEGs/TIBvsVH.csv", sep = ",", header = T)
DMEMvsVH <- read.csv("../Expression/DEGs/DMEMvsVH.csv", sep = ",", header = T)
TiPAvsVH <- read.csv("../Expression/DEGs/TIBPAvsVH.csv", sep = ",", header = T)
PAvsVH <- read.csv("../Expression/DEGs/PAvsVH.csv", sep = ",", header = T)
```

# Contrasting DEGs associated wih Tibolone

## Up-regulated


```{r listTi_up, message=FALSE}
devtools::install_github("gaospecial/ggVennDiagram", force = TRUE)

Tiup <- list(
  "TIBvsVH" = dplyr::filter(TivsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id ),
  "DMEMvsVH" = dplyr::filter(DMEMvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
Tiup
```

Now, we will build the Venn diagram

```{r Tivenn_up}
Tiup_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(Tiup)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(Tiup), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(Tiup)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(Tiup)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/TIB_up.pdf", plot = Tiup_venn, dpi = 600, width = 12, height = 8)
```


## Down-regulated

```{r listTi_down, message=FALSE}

Tidown <- list(
  "TIBvsVH" = dplyr::filter(TivsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id ),
  "DMEMvsVH" = dplyr::filter(DMEMvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
Tidown
```


```{r Tivenn_down}
Tidown_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(Tidown)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(Tidown), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(Tidown)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(Tidown)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/TIB_down.pdf", plot = Tidown_venn, dpi = 600, width = 12, height = 8)
```


# Check of DEGs associated wih Tibolone based on Tibolone + Palmitic Acid

## Up-regulated


```{r listTiPA_up, message=FALSE}

TiPAup <- list(
  "PAvsVH" = dplyr::filter(PAvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id ),
  "TIBPAvsVH" = dplyr::filter(TiPAvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id ),
  "TIBvsVH" = dplyr::filter(TivsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id ),
  "DMEMvsVH" = dplyr::filter(DMEMvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
TiPAup
```


```{r TiPAvenn_up}
TiPAup_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(TiPAup)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(TiPAup), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(TiPAup)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(TiPAup)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black", "Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/TIBPA_up.pdf", plot = TiPAup_venn, dpi = 600, width = 12.5, height = 8)

TiPAup@region[["item"]][[8]]
```


## Down-regulated

```{r listTiPA_down, message=FALSE}

TiPAdown <- list(
  "PAvsVH" = dplyr::filter(PAvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id ),
  "TiPAvsVH" = dplyr::filter(TiPAvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id ),
  "TivsVH" = dplyr::filter(TivsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id ),
  "DMEMvsVH" = dplyr::filter(DMEMvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
TiPAdown
```

```{r TiPAvenn_down}
TiPAdown_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(TiPAdown)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(TiPAdown), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(TiPAdown)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(TiPAdown)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black", "Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/TIBPA_down.pdf", plot = TiPAdown_venn, dpi = 600, width = 12.5, height = 8)
```


# DEGs associated wih Palmitic Acid

## Up-regulated


```{r listPA_up, message=FALSE}

PAup <- list(
  "PAvsVH" = dplyr::filter(PAvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id ),
  "TiPAvsVH" = dplyr::filter(TiPAvsVH, regulation == "Up") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
PAup
```

```{r PAvenn_up}
PAup_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(PAup)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(PAup), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(PAup)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(PAup)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/PA_up.pdf", plot = PAup_venn, dpi = 600, width = 12, height = 8)
```

```{r PA-interception_UP}

PAUp_TiPAvsVH <- TiPAvsVH %>%
  dplyr::filter(regulation == "Up") %>% 
                               dplyr::select( gene_id, symbol, biotype, logFC, PValue, FDR )
PAUp_PAvsVH <- PAvsVH %>%
  dplyr::filter(regulation == "Up") %>% 
                               dplyr::select( gene_id, symbol, biotype, logFC, PValue, FDR )

PA_common.up <- PAUp_TiPAvsVH %>%
    dplyr::rename( TiPAvsVH_symbol = symbol, TiPAvsVH_biotype = biotype,
                   TiPAvsVH_logFC = logFC, TiPAvsVH_PValue = PValue, TiPAvsVH_FDR = FDR ) %>%
    merge( PAUp_PAvsVH, ., by = "gene_id" ) %>%
    dplyr::rename( PAvsVH_symbol = symbol, PAvsVH_biotype = biotype,
                   PAvsVH_logFC = logFC, PAvsVH_PValue = PValue, PAvsVH_FDR = FDR ) %>%
    dplyr::select( -c( TiPAvsVH_symbol, TiPAvsVH_biotype ) ) %>% 
    dplyr::arrange( TiPAvsVH_FDR, PAvsVH_FDR )
   

PA_DEGup <- PAUp_PAvsVH %>%
  dplyr::rename( PAvsVH_symbol = symbol, PAvsVH_biotype = biotype,
                   PAvsVH_logFC = logFC, PAvsVH_PValue = PValue, PAvsVH_FDR = FDR ) %>% 
  dplyr::anti_join( PA_common.up, by = "gene_id" ) %>%
  dplyr::bind_rows( PA_common.up) %>% 
  dplyr::arrange( TiPAvsVH_FDR, PAvsVH_FDR )


write.table(PA_DEGup, file="../Overlapping/overlap_DEGs/PA-DEG_up.csv", quote = F, row.names = F, sep = ",")
head(PA_DEGup) 
```

## Down-regulated


```{r listPA_down, message=FALSE}

PAdown <- list(
  "PAvsVH" = dplyr::filter(PAvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id ),
  "TiPAvsVH" = dplyr::filter(TiPAvsVH, regulation == "Down") %>% 
                               dplyr::pull( gene_id )
             ) %>%
  RVenn::Venn() %>% 
  ggVennDiagram::process_data()
PAdown
```

```{r PAvenn_down}
PAdown_venn <- ggplot() +
  geom_sf(aes(fill = count), 
          data = ggVennDiagram::venn_region(PAdown)) +
  geom_sf(aes(color = id), size = 1,
          data = ggVennDiagram::venn_setedge(PAdown), show.legend = FALSE) +
  geom_sf_text(aes(label = name), size = 14, fontface = "bold", 
          data = ggVennDiagram::venn_setlabel(PAdown)) +
  geom_sf_text(aes(label = count), size = 14, 
          data = ggVennDiagram::venn_region(PAdown)) +
  ggplot2::scale_fill_gradient(low="HoneyDew", high = "LightBlue") +
  theme_void() +
  theme(legend.position = "none") +
  scale_colour_manual(values = c("Black", "Black")) +
  coord_sf(clip = "off")
  

ggsave("../Overlapping/venn/PA_down.pdf", plot = PAdown_venn, dpi = 600, width = 12, height = 8)
```

```{r PA-interception_DOWN}

PAUp_TiPAvsVH <- TiPAvsVH %>%
  dplyr::filter(regulation == "Down") %>% 
                               dplyr::select( gene_id, symbol, biotype, logFC, PValue, FDR )
PAUp_PAvsVH <- PAvsVH %>%
  dplyr::filter(regulation == "Down") %>% 
                               dplyr::select( gene_id, symbol, biotype, logFC, PValue, FDR )

PA_common.down <- PAUp_TiPAvsVH %>%
    dplyr::rename( TiPAvsVH_symbol = symbol, TiPAvsVH_biotype = biotype,
                   TiPAvsVH_logFC = logFC, TiPAvsVH_PValue = PValue, TiPAvsVH_FDR = FDR ) %>%
    merge( PAUp_PAvsVH, ., by = "gene_id" ) %>%
    dplyr::rename( PAvsVH_symbol = symbol, PAvsVH_biotype = biotype,
                   PAvsVH_logFC = logFC, PAvsVH_PValue = PValue, PAvsVH_FDR = FDR ) %>%
    dplyr::select( -c( TiPAvsVH_symbol, TiPAvsVH_biotype ) ) %>% 
    dplyr::arrange( TiPAvsVH_FDR, PAvsVH_FDR )
   

PA_DEGdown <- PAUp_PAvsVH %>%
  dplyr::rename( PAvsVH_symbol = symbol, PAvsVH_biotype = biotype,
                   PAvsVH_logFC = logFC, PAvsVH_PValue = PValue, PAvsVH_FDR = FDR ) %>% 
  dplyr::anti_join( PA_common.down, by = "gene_id" ) %>%
  dplyr::bind_rows( PA_common.down) %>% 
  dplyr::arrange( TiPAvsVH_FDR, PAvsVH_FDR )


write.table(PA_DEGdown, file="../Overlapping/overlap_DEGs/PA-DEG_down.csv", quote = F, row.names = F, sep = ",")
head(PA_DEGdown)  
```

## Session info

```{r session_information}
Sys.time()
getwd()
  
sessionInfo()
```