---
title: "Identification of hub genes for astrocytes"
output: html_notebook
author: Alexis Felipe Rojas Cruz
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.keep = 'all',
	dpi = 600,
	fig.width = 11,
	fig.height = 8.5
)
```



```{r include=FALSE, comment = ""}
if (!("clusterProfiler" %in% rownames(installed.packages()))){
    install.packages("clusterProfiler")
}
if (!("org.Hs.eg.db" %in% rownames(installed.packages()))){
    install.packages("org.Hs.eg.db")
}
if (!("ggVennDiagram" %in% rownames(installed.packages()))){
    install.packages("ggVennDiagram")
}
if (!("sf" %in% rownames(installed.packages()))){
    install.packages("sf")
}
if (!("RColorBrewer" %in% rownames(installed.packages()))){
    install.packages("RColorBrewer")
}
if (!("tidyverse" %in% rownames(installed.packages()))){
    install.packages("tidyverse")
}
if (!("ggplot2" %in% rownames(installed.packages()))){
    install.packages("ggplot2")
}
if (!("ggrepel" %in% rownames(installed.packages()))){
    install.packages("ggrepel")
}
if (!("devtools" %in% rownames(installed.packages()))){
    install.packages("devtools")
}
if (!("ggsci" %in% rownames(installed.packages()))){
    install.packages("ggsci")
}
if (!("ComplexUpset" %in% rownames(installed.packages()))){
    install.packages("ComplexUpset")
}
if (!("biomaRt" %in% rownames(installed.packages()))){
    install.packages("biomaRt")
}
BiocManager::install("STRINGdb")

library(clusterProfiler)
library(org.Hs.eg.db)
library(ggVennDiagram)
library(sf)
library(RColorBrewer)
library(tidyverse)
library(ggplot2)
library(devtools)
library(ggsci)
library(biomaRt)
library(ComplexUpset)
library(STRINGdb)
```

# Setting data for constructing PPI network 

### Loading data

For the construction of the PPI network, we will use the enriched DEGs that are found in the BP, CC and MF categories

```{r data}
PA <- read.csv("../Enrichment/enrich_GO/GO_PA.csv", sep = ",", header = T)
```

## Palmitic Acid

### Data Manipulation

```{r PA_preparing, message=FALSE, warning=FALSE}
PA_PPI <- PA %>% 
  dplyr::select( geneID ) %>% 
  tidyr::separate_rows( geneID ) %>% 
  dplyr::select( geneID ) %>% 
  dplyr::rename( hgnc_symbol = geneID ) %>% 
  unique( . )

mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", path="/biomart/martservice", dataset="hsapiens_gene_ensembl")
PA_ensembl <- getBM(values=PA_PPI$hgnc_symbol, filters= "hgnc_symbol", attributes= c( "hgnc_symbol", "gene_biotype"), mart= mart)

PA_proteins <- merge( PA_PPI, PA_ensembl, by = "hgnc_symbol") %>% 
  dplyr::filter( gene_biotype == "protein_coding") %>% 
  dplyr::select( hgnc_symbol ) %>%
  dplyr::arrange( hgnc_symbol )
```

Save data

```{r save}
write.csv(PA_proteins, file="../PPI_network/PA_proteins.csv", quote = F, row.names = F)
```

# Enrichment analysis of the modular genes

Cytoscape software was used to visualize the results. The modules of PPI were identified by MCODE tool.


### Loading data


```{r loading_data, modules}
module.1_raw <- read.csv("../PPI_network/modules/module_1/raw-module.csv", sep = ",", header = T)
module.2_raw <- read.csv("../PPI_network/modules/module_2/raw_module.csv", sep = ",", header = T)
```


## Module 1

```{r module_1, preparation}

module_1 <- module.1_raw %>%
  dplyr::select(string_id = X.id, symbol = display.name, string_name = stringdb..canonical.name, string_sequece = stringdb..sequence,
                string_description = stringdb..description, MCODE_status = MCODE..Node.Status..1., MCODE_score =      MCODE..Score..1.) %>%
  dplyr::arrange( desc( MCODE_score ) )

head(module_1)

write.table(module_1, file="../PPI_network/modules/module_1/curated-module.csv", quote = F, row.names = F, sep = "\t")
```

### Enrichment

```{r module_1, GO}
module_1.genes <- clusterProfiler::bitr(module_1$symbol, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db" ) %>%
  dplyr::select( ENTREZID ) %>%
  dplyr::rename( Entrez = ENTREZID )

module_1.GO_BP = enrichGO(module_1.genes$Entrez, ont = "BP", OrgDb = org.Hs.eg.db, pAdjustMethod = "BH", pvalueCutoff  = 0.01,
                           qvalueCutoff  = 0.05, readable = TRUE)
                           

module1_GO <- dotplot(module_1.GO_BP, showCategory = 10) +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 18),
  axis.title.x = element_text(size = 24), panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22),
  plot.title = element_text(size = 28, hjust = 0.5)) +
  ggtitle("Biological Process")
module1_GO

ggsave("../PPI_network/modules/module_1/enrichment/GO/module1_GO.pdf", plot = module1_GO, dpi = 600, width = 9, height = 8)
```

```{r module_1, KEGG}

module_1.KEGG = enrichKEGG(module_1.genes$Entrez, organism = "hsa", pAdjustMethod = 'BH', pvalueCutoff = 0.01,
                            qvalueCutoff  = 0.05)
                           

module1_KEGG <- dotplot(module_1.KEGG, showCategory = 10) +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 18),
  axis.title.x = element_text(size = 24), panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22),
  plot.title = element_text(size = 28, hjust = 0.5)) +
  ggtitle("KEGG Pathway")
module1_KEGG

ggsave("../PPI_network/modules/module_1/enrichment/KEGG/module1_KEGG.pdf", plot = module1_KEGG, dpi = 600, width = 9, height = 8)
```


### Save data

```{r module_1, save}

module1_enrichGO <- module_1.GO_BP@result
module1_enrichKEGG <- module_1.KEGG@result

write.table(module1_enrichGO, file="../PPI_network/modules/module_1/enrichment/GO/module1_GO.csv", quote = F, row.names = F, sep = ";")
write.csv(module1_enrichKEGG, file="../PPI_network/modules/module_1/enrichment/KEGG/module1_KEGG.csv", quote = F, row.names = F)

```

## Module 2

```{r module_2, preparation}

module_2 <- module.2_raw %>%
  dplyr::select(string_id = X.id, symbol = display.name, string_name = stringdb..canonical.name, string_sequece = stringdb..sequence,
                string_description = stringdb..description, MCODE_status = MCODE..Node.Status..1., MCODE_score =      MCODE..Score..1.) %>%
  dplyr::arrange( desc( MCODE_score ) )

head(module_2)

write.table(module_2, file="../PPI_network/modules/module_2/curated-module.csv", quote = F, row.names = F, sep = "\t")
```

### Enrichment

```{r module_2, GO}
module_2.genes <- clusterProfiler::bitr(module_2$symbol, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db" ) %>%
  dplyr::select( ENTREZID ) %>%
  dplyr::rename( Entrez = ENTREZID )

module_2.GO_BP = enrichGO(module_2.genes$Entrez, ont = "BP", OrgDb = org.Hs.eg.db, pAdjustMethod = "BH", pvalueCutoff  = 0.01,
                           qvalueCutoff  = 0.05, readable = TRUE)
                           

module2_GO <- dotplot(module_2.GO_BP, showCategory = 10) +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 18),
  axis.title.x = element_text(size = 24), panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22),
  plot.title = element_text(size = 28, hjust = 0.5)) +
  ggtitle("Biological Process")
module2_GO

ggsave("../PPI_network/modules/module_2/enrichment/GO/module2_GO.pdf", plot = module2_GO, dpi = 600, width = 9, height = 8)
```

```{r module_2, KEGG}

module_2.KEGG = enrichKEGG(module_2.genes$Entrez, organism = "hsa", pAdjustMethod = 'BH', pvalueCutoff = 0.01,
                            qvalueCutoff  = 0.05)
                           

module2_KEGG <- dotplot(module_2.KEGG, showCategory = 10) +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 18),
  axis.title.x = element_text(size = 24), panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22),
  plot.title = element_text(size = 28, hjust = 0.5)) +
  ggtitle("KEGG Pathway")
module2_KEGG

ggsave("../PPI_network/modules/module_2/enrichment/KEGG/module2_KEGG.pdf", plot = module2_KEGG, dpi = 600, width = 9, height = 8)
```


### Save data

```{r module_2, save}

module2_enrichGO <- module_2.GO_BP@result
module2_enrichKEGG <- module_2.KEGG@result

write.table(module2_enrichGO, file="../PPI_network/modules/module_2/enrichment/GO/module2_GO.csv", quote = F, row.names = F, sep = ";")
write.csv(module2_enrichKEGG, file="../PPI_network/modules/module_2/enrichment/KEGG/module2_KEGG.csv", quote = F, row.names = F)

```


# Identification of hub genes

The hub genes were screened out by the cytoHubba plugin, a tool that identifies hub genes through multiple algorithms. The intersections of four ranking algorithms, including Maximal Clique Centrality (MCC), Density of Maximum Neighborhood Component (DMNC), Maximum Neighborhood Component (MNC), Degree, were considered as hub genes.

### Loading data


```{r loading_data, hub_genes}
MCC_raw <- read.csv("../PPI_network/hub_genes/MCC_top20.csv", sep = ",", header = T)
DMNC_raw <- read.csv("../PPI_network/hub_genes/DMNC_top20.csv", sep = ",", header = T)
MNC_raw <- read.csv("../PPI_network/hub_genes/MNC_top20.csv", sep = ",", header = T)
Degree_raw <- read.csv("../PPI_network/hub_genes/Degree_top20.csv", sep = ",", header = T)
```


### Setting the files


```{r hub genes, preparing_data}
string_db <- STRINGdb$new( version="11.5", species=9606, score_threshold=400)

term <- string_db$get_proteins()

#MCC 
MCC <- MCC_raw %>%
  dplyr::select( Name, Score ) %>% 
  dplyr::rename( protein_external_id = Name ) %>% 
  merge( term, ., by = "protein_external_id" ) %>%
  dplyr::select( protein_external_id, preferred_name, Score ) %>%
  dplyr::mutate( protein_external_id = str_remove_all(protein_external_id, "9606." )) %>% 
  dplyr::rename( Protein_ID = protein_external_id, Symbol = preferred_name ) %>%
  tibble::add_column(Method = "MCC", .before = "Protein_ID") %>% 
  dplyr::arrange( desc(Score) )

#DMNC 
DMNC <- DMNC_raw %>%
  dplyr::select( Name, Score ) %>% 
  dplyr::rename( protein_external_id = Name ) %>% 
  merge( term, ., by = "protein_external_id" ) %>%
  dplyr::select( protein_external_id, preferred_name, Score ) %>%
  dplyr::mutate( protein_external_id = str_remove_all(protein_external_id, "9606." )) %>% 
  dplyr::rename( Protein_ID = protein_external_id, Symbol = preferred_name ) %>%
  tibble::add_column(Method = "DMNC", .before = "Protein_ID") %>% 
  dplyr::arrange( desc(Score) )

#MNC
MNC <- MNC_raw %>%
  dplyr::select( Name, Score ) %>% 
  dplyr::rename( protein_external_id = Name ) %>% 
  merge( term, ., by = "protein_external_id" ) %>%
  dplyr::select( protein_external_id, preferred_name, Score ) %>%
  dplyr::mutate( protein_external_id = str_remove_all(protein_external_id, "9606." )) %>% 
  dplyr::rename( Protein_ID = protein_external_id, Symbol = preferred_name ) %>%
  tibble::add_column(Method = "MNC", .before = "Protein_ID") %>% 
  dplyr::arrange( desc(Score) )

#Degree
Degree <- Degree_raw %>%
  dplyr::select( Name, Score ) %>% 
  dplyr::rename( protein_external_id = Name ) %>% 
  merge( term, ., by = "protein_external_id" ) %>%
  dplyr::select( protein_external_id, preferred_name, Score ) %>%
  dplyr::mutate( protein_external_id = str_remove_all(protein_external_id, "9606." )) %>% 
  dplyr::rename( Protein_ID = protein_external_id, Symbol = preferred_name ) %>%
  tibble::add_column(Method = "Degree", .before = "Protein_ID") %>% 
  dplyr::arrange( desc(Score) )
```

```{r hub_genes save}
write.csv(MCC, file="../PPI_network/hub_genes/curated-hub/MCC.csv", quote = F, row.names = F)
write.csv(MNC, file="../PPI_network/hub_genes/curated-hub/MNC.csv", quote = F, row.names = F)
write.csv(DMNC, file="../PPI_network/hub_genes/curated-hub/DMNC.csv", quote = F, row.names = F)
write.csv(Degree, file="../PPI_network/hub_genes/curated-hub/Degree.csv", quote = F, row.names = F)
```

### Upset diagram

```{r hub genes, upset}

hub_genes <- data.frame(genes = c(MCC$Symbol, MNC$Symbol, DMNC$Symbol, Degree$Symbol),
                        method = c(MCC$Method, MNC$Method, DMNC$Method, Degree$Method)) %>%
  dplyr::count(genes, method) %>% 
  pivot_wider(names_from = method,
              values_from = n,
              values_fill = list(n = 0))

method = colnames(hub_genes)[2:5]

p1 <- upset(
    hub_genes,
    method,
    base_annotations=list(
        'Intersection size'=intersection_size(
            mapping=aes(fill='bars_color')
        ) + scale_fill_manual(values=c('bars_color'='PaleGoldenrod'), guide='none')
    ),
    encode_sets=FALSE,
    matrix=(
        intersection_matrix(
            geom=geom_point(
                shape='square',
                size=5
            ),
            segment=geom_segment(
                linetype='dotted'
            ),
            outline_color=list(
                active='darkorange3',
                inactive='grey70'
            )
        )
        + scale_color_manual(
            values=c('TRUE'='orange', 'FALSE'='grey'
            )
        )
        + scale_y_discrete(
            position='right'
        )
    ),
    queries=list(
        upset_query(
            intersect=c('MNC', 'Degree', 'MCC', 'DMNC'),
            color='DarkSeaGreen',
            fill='DarkSeaGreen',
            only_components=c('intersections_matrix', 'Intersection size')
        )
    )
)

ggsave("../PPI_network//hub_genes/common-hub/Upset.pdf", plot = p1, dpi = 600, width = 8, height = 6)
```


Retrieve the common hub genes

```{r hub genes common}

hub_common <- Degree %>%
    dplyr::select(-Method) %>% 
    dplyr::rename( Degree_Score = Score ) %>%
    merge( DMNC, ., by = c("Symbol", "Protein_ID") ) %>%
    dplyr::rename( DMNC_Score = Score ) %>%
    dplyr::select(-Method) %>% 
    merge( MNC, ., by = c("Symbol", "Protein_ID") ) %>%
    dplyr::rename( MNC_Score = Score ) %>%
    dplyr::select(-Method) %>% 
    merge( MCC, ., by = c("Symbol", "Protein_ID") ) %>%
    dplyr::rename( MCC_Score = Score ) %>%
    dplyr::select(-Method) %>% 
    dplyr::arrange( desc( MNC_Score ) )

head(hub_common)
```

```{r Ti_commom hub_genes save}
write.csv(hub_common, file="../PPI_network/hub_genes/common-hub/common.csv", quote = F, row.names = F)
```

### Enrichment


```{r hub_genes, GO}
hub <- clusterProfiler::bitr(hub_common$Symbol, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db" ) %>%
  dplyr::select( ENTREZID ) %>%
  dplyr::rename( Entrez = ENTREZID )

hub_GO = enrichGO(hub$Entrez, ont = "BP", OrgDb = org.Hs.eg.db, pAdjustMethod = "BH", pvalueCutoff  = 0.01,
                           qvalueCutoff  = 0.05, readable = TRUE)
                           

hub.common_GO <- dotplot(hub_GO, showCategory = 10) +
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#371ea3"),
  guide=guide_colorbar(reverse=TRUE, order=1)) +
  guides(size = guide_legend(override.aes=list(shape=1))) +
  theme(panel.grid.major.y = element_line(linetype='dotted', color='#808080'),
  axis.text.y = element_text(size = 18), axis.text.x = element_text(size = 18),
  axis.title.x = element_text(size = 24), panel.grid.major.x = element_blank()) +
  theme(strip.text.x = element_text(size = 24), strip.text.y = element_text(size = 24),
  legend.title = element_text(size = 26, face = "bold"), legend.text = element_text(size = 22),
  plot.title = element_text(size = 28, hjust = 0.5)) +
  ggtitle("Biological Process")
hub.common_GO

ggsave("../PPI_network/hub_genes/common-hub//enrichment/GO/GO_hub.pdf", plot = hub.common_GO, dpi = 600, width = 9, height = 8)
```

```{r hub_genes, save}

hub_genes_GO <- hub_GO@result

write.table(hub_genes_GO, file="../PPI_network/hub_genes/common-hub//enrichment/GO/GO_hub.csv", quote = F, row.names = F, sep = ";")

```


## Session info

```{r session_information}
Sys.time()
getwd()
  
sessionInfo()
```